---
title: "PGA Tour Project"
author: "Caleb Fornshell"
date: "11/11/2020"
output: html_document
---


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(tidyverse)
PGAdata=read_csv("C:\\Users\\Caleb\\Desktop\\jobs\\Projects\\PGA Tour\\PGA_cleaned.csv")
```

## Objective:
In this project I want to analyze data that is based on PGA Tour players season
statistics. I want to perform cluster analysis and determine which golfers could 
be considered similar to one another. I will also perform use regression 
techniques to generate models for predicting a golfers probability of winning
in a particular year, and a golfers stroke average.

## Data Collection and Cleaning
This data was collected from the official PGA Tour website. The data was scraped
using Power Query in Excel. Each statistic used had to be scraped to its own 
table The tables were then read into SAS and merged using PROC SQL. Once the 
preprocessing of data was completed it was read into R.

## Variables
In total the dataset contains 18 variables, of which 3 are identifying variables
2 are response variables, and 12 are predictor variables. A complete description 
of the variables can be found here. A subset of the data is shown below

```{r}
summary(str(PGAdata))
PGAdata[sample(1:100,6,replace=FALSE),]
PGAdata
```



## Exploratory Data Analysis

### Golfers and Years Analysed
```{r}
length(unique(PGAdata$Golfer))
unique(PGAdata$Year)
```

