---
title: "Driving Distance Project"
output: 
  html_document:
    theme: cosmo
    code_folding: "hide"
    df_print: paged
    toc: TRUE
    toc_depth: "1"
---

<!-- 
This file creates the the html document for my Driving Distance Project.
The data for this project can be found in the driving_master.csv file.

Objective of the project is to determine whether driving distance has seen a 
statistically significant increase in driving distance and what caused the 
change if a change is found.
-->


<!-- increasing text size for whole document -->
<style type="text/css">

body, td {
   font-size: 18px;
}
</style>

<!-- Styling for blue banner -->
<style>
div.blue { background-color:#008af4; 
           border-radius: 5px; 
           padding: 10px;
           color: white;
           }
</style>

<!-- Styling for gray box around tables and graphs -->
<style>
div.back { background-color: white;
  border: 3px solid lightgray;
  padding: 3px;
  margin: 20px;
           }
</style>


<!-- Installing packages and reading in data -->
```{r setup, include=FALSE}
require(tidyverse)
require(reshape2) #melt() function
require(plotly) #interactive graphics
require(DT) #datatables

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
infile="C:\\Users\\Caleb\\Desktop\\jobs\\Projects\\Driving Distance\\data\\Driving_Master.csv"


Drive = read.csv(infile, header = TRUE, encoding = "UTF-8")
```

<div  class="blue">
# Introduction
</div>


The purpose of this study is to analyze tee shots on the PGA tour, and discuss 
how these shots have distances have changed during the 2007-2019 seasons. 
Some of the research questions examined in this analyse include:

  + Has Driving Distance increased and by how much?
  + What variables are correlated with increased distance?
  + Are other correlations noteworthy?
  + Has the amount of roll changed over time?
  
Other than `DrivingDistance`, the variables discussed in this study are based on the 'Off the Tee' statistics
that are captured using radar as described by the PGA Tour's [website](https://www.pgatour.com/stats/categories.ROTT_INQ.html). The data was scraped from the PGA Tour's website using a web scraper that created a table
for each statistic. These tables were then loaded into a mySQL database and 
joined to create a single table. This merged table was used in the analysis 
below. Links to the web scraper and SQL script are found [here](https://github.com/CalebF94/Data-Analysis-Projects/tree/main/Driving%20Distance%20Project). 

A note on notation: A variable's name is `highlighted` if the reference is 
specific to a column in the data set.

<br>
<div  class="blue">
# Exploratory Data Analysis
</div>

## *Structure of the Data* {.tabset .tabset-pills}

******

The data contains 22 variables,of which 9 are statisitics of interest and 9 are 'Attempt' variables indicating the number of measurements that were taken for a particular statistic. The dataset contains 2437 observations and 540 unique 
golfers. Each row represents a single season for a particular golfer. The data 
also contains no missing information. The dataset(with 'Attempt' variables 
removed) is shown below.



### Data
<div  class="back">
```{r}
#using DT package to print only numeric variable columns
datatable(Drive[,c(2:4,seq(5,21,2))], 
          options = list(scrollX=TRUE,
                         pageLength=8))

```
</div>

### Observations
<div  class="back">
```{r}
obs = dim(Drive)[1]
vars =  dim(Drive)[2]

#First and last year
y1 = min(Drive$Year)
y2 = max(Drive$Year)

#giving basic counts for dataset
tibble("Number of Obs" = obs,
       "Number of Vars" = vars,
       "Unique Golfers" = length(unique(Drive$Player)))
```
</div>

### Missing Data
<div  class="back">
```{r}
#determing which variables, if any, have missing values
apply(apply(Drive, 2, is.na), 2, sum)#no missing values
```
</div>

## {-}



## *Descriptive Statistics* {.tabset .tabset-pills}

********


The following tabs contain basic counts and descriptive statistics for all
variables. The "Year" tab shows that on average 187 golfers participated each year,
and that the number of golfers each season ranged from 177-195. The "Table"
tab contains the sample mean, median, standard deviation, minimum and maximum
for each variable along with the units of measurement.


### Year 
<div  class="back">
```{r  fig.height=3}
# golfers in each year
# table(Drive$Year)
ggplot(data=Drive, aes(x=as.factor(Year))) + 
  geom_bar() +
  coord_flip()+
  geom_text(stat='count', aes(label=..count..), 
            hjust=-.1, position =position_dodge(width=.8)) +
  xlab("Year") +
  labs(title = "PGA Golfers by Year", 
       subtitle = paste("Average: ", round(obs/length(unique(Drive$Year)))))+
  theme(axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x  = element_blank())


```
</div>

### Table
<div  class="back">
```{r}

#selecting only numeric columns, and getting unit labels
numericvars=Drive %>% select(seq(5,21,2))
label=c("Yds", "Ft", "Yds", "MPH", "Yds", "MPH", "Sec", "Deg", "RPM")

#creating descriptive statistics table
Descr.Table = 
    tibble(Variable = paste(names(numericvars), " (", label, ")", sep=""),
           Count = apply(numericvars,2 , length),
           Average = round(apply(numericvars, 2, mean),2),
           Median = apply(numericvars, 2, median),
           Std.Dev = apply(numericvars, 2, sd),
           Minimum = apply(numericvars, 2, min),
           Maximum = apply(numericvars, 2, max))

Descr.Table
```
</div>
## {-}

## *Distribution of Variables* {.tabset .tabset-pills}

*******

The distribution of each variable is shown using histograms, density plots, and
boxplots in the following tabs. These graphs show that the variables are 
symmetric and there doesn't appear to be any obvious major issues. The boxplots
do show that there are a few outliers for each variable, but that is not unusual
for the number of observations being studied.

### Density Plots
<div  class="back">
```{r message=FALSE, fig.align='center'}
# Distribution of numeric variables
ggplot(data = melt(Drive %>% select(5,7,9,11,13,15,17,19))) +
  geom_histogram(aes(x=value, y=..density..), bins = 20) +
  geom_density(aes(x=value, color = variable), size=1.2) +
  facet_wrap(~variable, scales = "free", ncol = 4) +
  guides(color=FALSE) +
  labs(title = "Histograms and Density Plots of Numeric Variables")
```
</div>

### Boxplots 
<div  class="back">
```{r fig.align='center'}

# boxplots for each numeric variable
ggplot(data = melt(Drive %>% select(3,5,7,9,11,13,15,17,19,21), id.vars = 1)) +
  geom_boxplot(aes(x=value, y=variable, color=variable)) +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank()) +
  ylab("") + xlab("") +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  guides(color=FALSE) +
  labs(title = "Boxplots of Numeric Variables")

```
</div>
## {-}

## *Relationships Between Variables* {.tabset .tabset-pills}

*******

In order to determine which variables may be responsible for the possible increase
in driving distance, the correlation between variables is needed. The graphs
below illustrate which variables are most correlated. The most interesting 
relationships are shown and discussed further.

### Correlation Matrix

> I'm most interested in the `DrivingDistance` variable and it appears that
> `BallSpeed` and `CarryDistance` are the most highly correlated variables, which
> makes intuitive sense. `ClubheadSpeed` and `ApexDistance` also have very high
> correlations. Other notable correlations exist between `LaunchAngle` and 
>`ClubheadSpeed` and `BallSpeed`.

<div  class="back">
```{r}
# correlation matrix
corrplot::corrplot(cor(numericvars), method = "number", type = "lower",
                   addCoef.col = FALSE, tl.srt = 15, diag = FALSE)

```
</div>

### Driving Distance Scatterplots

> These relationships all make sense intuitively, and are really not at all 
> surprising. If modeling driving distance was a goal, the high correlation
> between these variables would need to be accounted for.

<div  class="back">
```{r}
# scatter plots with variables highly correlated to DrivingDistance
# Vars: Apex dist, ballspeed, carrydist, CHspeed
melt(Drive %>% select(5,9,11,13,15), id.vars = 1) %>%
ggplot(aes(x = DrivingDistance, y = value)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~variable, scales = "free", strip.position = "left") +
  ylab("") +
  labs(title = "Driving Distance Vs Highly Correlated Variables")
```
</div>

### Launch Angle Scatterplots

> I also found these correlations interesting as they were the only significant
> negative correlations, and there does not seem to be an intuitive reason why
> these correlations are negative.

<div  class="back">
```{r}
# Scatterplots for strong negative correlations
# strong negative correlations associated with Launch Angle
# Vars LaunchAngle, CHSpeed, BallSpeed, SpinRate
melt(Drive %>% select(19, 15, 11, 21), id.vars = 1) %>%
ggplot(aes(x=LaunchAngle, y=value)) +
  geom_point() +
  facet_wrap(~variable, scales = "free", strip.position = "left", ncol = 2) +
  geom_smooth(method = "lm") +
  ylab("") +
  labs(title = "Launch Angle Vs Negative Correlated Variables")
```
</div>
## {-}


<br>
<div  class="blue">
# Is Driving Distance Increasing? 
</div>

## *Trends Over Time*

*********

The plot below shows how the distributions of each variable have changed over 
time. Most variables appear to have increasing trendlines, including all 
distance related variables. This seems to suggest that driving distance has 
indeed increased over the last 13 years. Later, a statistical test will be used 
to determine whether the increase is significant. Three variables, `HangTime`, `LaunchAngle`, and `SpinRate`, appear either be non-increasing or possibly
decreasing.

<div  class="back">
```{r fig.width=8}
# change of all variables over time
#LA and spin rate decrease
melt(Drive %>% select(1, 2,5,7,9,11,13,15,17,19,21), 
     id.vars = c("Year","PlayerYear")) %>%
ggplot(aes(x=as.factor(Year), y=value, label=PlayerYear)) +
  geom_boxplot() +
  geom_smooth(aes(group=1), method="lm") + 
  theme(axis.text.x = element_text(angle = 90))+
  facet_wrap(~variable, scales = "free_y", ncol=3, strip.position = "left") +
  ylab("") + xlab("") + 
  labs(title = "Variable Trends Over Time")
 

```
</div>
<br>

### *Is The Increase Significant?*{.tabset .tabset-pills}

********

To determine whether the average driving distance is increasing a simple linear
regression model is created with year as the predictor variable. The mean 
driving distance for each season was calculated using the driving distance 
weighted by the driving attempt variable. 

The results of the regression analysis show that the increase is significant as 
the p-value < 0.05. This model states that for every year after 2007, average 
driving distance increases on average by approximately 0.59 yards.

#### Graph
<div  class="back">
```{r}
# is average drive distance increasing over time?

# subsetting and aggregating data for Driving Distance
DD = 
  Drive %>%
  transmute(
    Year,
    Attempts,
    totaldistance = DrivingDistance*Attempts) %>% 
  group_by(Year) %>%
  summarize(AvgDrive=sum(totaldistance)/sum(Attempts))


#generates ANOVA table
m=summary(lm(AvgDrive~Year, DD))
 
#creating text for subtitle          
coeffs=round(m$coefficients,4)
form=paste(coeffs[1], " + ", coeffs[2],"*Year", sep="")

#scatterplot with regression line 
ggplot(data = DD, aes(x=Year, y=AvgDrive)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Linear Relationship Between Year and Average Driving Distance",
     subtitle = paste("Formula: DrivingDistance(yds) = ", form)) +
  scale_x_continuous(breaks = y1:y2)


```
</div>

#### Table
<div  class="back">
```{r}
#outputting ANOVA table
m
```

</div>


### *Is Average Carry Increasing?*{.tabset .tabset-pills}

*******

Driving distance is composed of two parts: carry distance and roll distance. I 
want to determine whether drives are going further due to carry distance or roll
distance. Again, simple linear regression models were used to quantify the 
relationship between carry distance and time.

This model suggests that carry distance is also increasing over time. This time 
the coefficent for the `Year` variable is 1.0813. This is means that on average 
carry distance increased by 1.0813 yards per season.

#### Graph
<div  class="back">
```{r}
# is average carry increasing?

# subsetting and aggregating data for Carry Distance
CD =
  Drive %>%
  transmute(
    Year,
    CDAttempts,
    totaldistance = CarryDistance*CDAttempts) %>% 
  group_by(Year) %>%
  summarize(AvgCarry=sum(totaldistance)/sum(CDAttempts))

#For ANOVA table
m=summary(lm(AvgCarry~Year, CD))

#for subtitle
coeffs=round(m$coefficients,4)
form=paste(coeffs[1], " + ", coeffs[2],"*Year", sep="")

#Scatterplot and regression line 
ggplot(data = CD, aes(x=Year, y=AvgCarry)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "Linear Relationship Between Year and Average Carry Distance",
       subtitle = paste("Formula: CarryDistance(yds) = ", form)) +
  scale_x_continuous(breaks = y1:y2)


```
</div>

#### Table
<div  class="back">
```{r}
#Outputting ANOVA table
m
```
</div>
<br>


### *What About Roll* {.tabset .tabset-pills}

*****

Roll was calculated simply by subtracting `CarryDistance` from `DrivingDistance`.
A regression model using this new roll variable suggests that the average roll
has decreased since 2007. The coefficient value for the `Year` variable suggests
that the average roll has decreased by 0.4941 yards per year on average. This result is not
surprising given the results from the `CarryDistance` regression. In fact, 
summing the `Year` coefficients for the roll and carry distance regressions is
equal to the coefficient value from the `DrivingDistance` regression (i.e. 1.0813 - 0.4941 = 0.5873).

#### Graph
<div  class="back">
```{r}
#What about Roll Distance?

#Joining CD and DD and using difference to calculate roll for each season
Roll = inner_join(CD, DD, by = "Year") %>%
  mutate(AvgRoll = AvgDrive - AvgCarry)

#For ANOVA table 
m=summary(lm(AvgRoll~Year, Roll))

#for subtitle
coeffs=round(m$coefficients,4)
form=paste(coeffs[1], " + ", coeffs[2],"*Year", sep="")

#Scatterplots and regression line for Roll Distance
Roll %>%
  ggplot(aes(x=Year, y=AvgRoll)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "Linear Relationship Between Year and Average Roll",
       subtitle = paste("Formula: Roll(yds) = ", form)) +
  scale_x_continuous(breaks = y1:y2) 



```
</div>


#### Table
<div  class="back">
```{r}
#ANOVA table
m
```
</div>

### {-}

The previous models show that overall driving distance has increased due to an
increase in carry distance even though the amount of roll has decreased. The 
decrease in roll is the only surprising result. Why has roll decreased even 
though distance, clubhead speed, and ball speed have not decreased? Two theories
are that the small overall decrease in spin rate reduces the amount of roll, and
that the increasing apex height results in a ball flight less condusive to roll.
These two theories are explored further.

<br>

## Why Has Average Roll Decreased? {.tabset .tabset-pills}

The following scatterplots show that there does appear to be an inverse
relationship between apex height and roll distance while there appears to be no
relationship between spin rate and roll distance. Based on this result alone, 
spin rate does not appear to explain the decrease in roll.

### Apex Height
<div  class="back">
```{r}
#Scatterplot for Average roll vs apex height
Drive %>%
  mutate(AvgRoll= DrivingDistance - CarryDistance,
         AOD = (180/pi)*atan(ApexHeight/(CarryDistance - ApexDistance))) %>%
  ggplot(aes(x=AvgRoll, y=ApexHeight) ) +
  geom_smooth(method = "lm", se=FALSE, color="black") +
  geom_point(color="#008af4")
```
</div>

### Spin Rate
<div  class="back">
```{r}
#scatterplot for average roll vs spin rate, no correlation
Drive %>%
  mutate(AvgRoll= DrivingDistance - CarryDistance,
         AOD = (180/pi)*atan(ApexHeight/(CarryDistance - ApexDistance))) %>%
  ggplot(aes(x=AvgRoll, y=SpinRate) ) +
  geom_smooth(method = "lm", se=FALSE, color="black") +
  geom_point(color="#008af4") 
```
</div>

## {-}

My theory to explain the negative relationship between apex height and roll is 
that the increase in height results in a more vertical descent of the ball and 
makes the shot less condusive to rolling. To test this theory, I calculated a 
variable called `AOD`, or Angle of Descent, which uses `ApexHeight`,
`CarryDistance`, `ApexDistance` and the arctangent function to calculate the 
angle between the apex and the landing spot. The graphic below crudely shows how
the angle of descent has changed over time. While the graph is not a perfect 
depiction of a shot, it does show that in general the angle of descent has 
increased.

<div  class="back">
```{r}
fp1 = # Data for Flight Path graphic
  Drive %>% 
  mutate(totalApexDist = ApexDistance * ADAttempts,
         TotalApexHeight = ApexHeight*AhAttempts/3) %>% #converted to yards
  group_by(Year) %>%
  #getting yearly summary
  summarize(AvgApexDistance = sum(totalApexDist)/sum(ADAttempts),
            AvgApexHeight = sum(TotalApexHeight)/sum(AhAttempts)) %>%
  inner_join(Roll, by = "Year") %>%
  # angle of descent: arctan(opposite/adj)
  mutate(AngleofDesc. = (180/pi)*atan(AvgApexHeight/(AvgCarry - AvgApexDistance)))

#format the data into a tibble for ggplot
fp = tibble(
            Year = as.factor(rep(unique(fp1$Year),3)),
            x = c(rep(0,13), fp1$AvgApexDistance, fp1$AvgCarry),
            y = c(rep(0,13), fp1$AvgApexHeight, rep(0,13)),
            AngleOfDescent = rep(round(fp1$AngleofDesc.,2),3))


# flight path graphic
ggplotly(
  tooltip=c("group", "colour"),
  ggplot(data=fp, aes(x=x, y=y, group=AngleOfDescent)) + 
  geom_path(aes(colour=Year)) +
  xlab("Distance(Yds)") + ylab("Height(Yds)") +
  labs(title = "Shot Shape and Angle of Descent")
    
)
```
</div>

## {.tabset .tabset-pills}
A regression model was made to quantify the relationship between angle of 
descent and time. Again, we see a significant increase in the angle of descent 
over time. This means the ball appears to be landing at a steeper angle which 
may be the reason behind the decrease in roll.

### Graph
<div  class="back">
```{r}
#Getting coefficients and formula for AOD regression line
m=summary(lm(AngleofDesc.~Year, data=fp1))
coeffs = round(m$coefficients,4)
form=paste(coeffs[1], " + ", coeffs[2],"*Year", sep="")

#scatterplot and regression line for AOD 
ggplot(fp1, aes(Year, AngleofDesc.)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "Linear Relationship Between AOD and Year",
       subtitle = paste("Formula: AOD =", form))
```
</div>

### Table
<div  class="back">
```{r}
m
```
</div>

## {-}

<br>
<div  class="blue">
# Conclusions 
</div>

This study supports the notion that driving distance has increased in recent
years on the PGA Tours. This increase appears to be largely correlated to the
increase in clubhead and ball speed. In addition, it appears that the increase
is due to ball travelling farther in the air as the distance the ball rolls has
decreased. A possible explanation why drives are not rolling as far my be that
there has been a increase in the angle of descent and drives are coming into 
the ground on a steeper angle.

This study may be improved data was unaggregated. The data originally 
came as season long statistics. The ability to look at each stroke individually
may allow for a better understanding of the relationships between the variables.
Additionally, information on which club was used may be useful. According to the 
PGA website this information was collected on par 4 and 5 holes, but it is not 
guaranteed that a driver was used. This may slightly influence the results of 
this analysis if the usage of driver has changed over time. 

